---
title: 'Hagelloch: Exploratory Analysis'
author: "Matthew Smith"
date: "16/11/2020"
output: html_document
---

## Setup

Load required libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("surveillance")
library(mgcv)
library(numDeriv)
library(gdata)
```

Import data and set x.loc and y.loc as coordinates (for use later).

```{r include=FALSE}
data("hagelloch")
hagelloch = hagelloch.df
coordinates(hagelloch.df) <- c("x.loc","y.loc")
```


## SIR model

We begin with a simple SIR model assuming the infection and recovery times in the dataset are accurate.

```{r SIR MODEL, echo=FALSE}
n = length(hagelloch$PN)

days = 0:93
obs.I.SIR = vector()
obs.R.SIR = vector()


for (i in 1:length(days)){
  X = 0
  Y = 0
  for (j in 1:n){
    if (hagelloch$tI[j] <= days[i] && days[i] <= hagelloch$tR[j]){
      X = X + 1
    }
    
    if (days[i] >= hagelloch$tR[j]){
      Y = Y + 1
    }
  }
  obs.I.SIR = c(obs.I.SIR, X)
  obs.R.SIR = c(obs.R.SIR, Y)
}



obs.S.SIR <- 188 - obs.I.SIR - obs.R.SIR
#plot.new(removed)
#plot(removed, type = "l")
plot(obs.S.SIR, type = "l", col = "green", xlab = "Time (days)", ylab = "Number of individuals", main = "Infectious states of individuals over time")
lines(obs.I.SIR, col = "red")
lines(obs.R.SIR, col = "blue")
abline(h=0, lty = 2)
abline(h=188, lty = 2)
grid()
legend(70, 125, legend=c("susceptible", "infectious", "removed"), col = c("green", "red", "blue"), lty = c(1, 1), cex = 0.9)
```

## SEIR model

We now move onto an SEIR model where we assume:

- A latent period of 9 days from exposure to the beginning of the infectious period
- A one day period of from the infectious period beginning until prodoromal symptoms begin to show
- A 3 day period after the onset of rash until recovery.

```{r SEIR model, echo=FALSE}

obs.SEIR_plot = function(function_data, e, k, d, SEIR.plot = TRUE){

  # Number of individuals
  
  n = length(function_data$PN)
  
  # Day range of whole epidemic
  
  days = 0:ceiling(max(function_data$tERU + d))
  
  # Initialise vectors
  
  obs.E.SEIR = vector()
  obs.I.SEIR = vector()
  obs.R.SEIR = vector()
  obs.E_added = vector()
  
  
  for (i in 1:length(days)){
  
    X = 0     # Number exposed on day i
    Y = 0     # Number infectious on day i
    Z = 0     # Number removed on day i
    S = 0     # Added exposed on day i
    
    for (j in 1:n){
      
      # Find actual exposure, infectious and removed times based on our assumptions
      
      exposure_time <- function_data$tPRO[j] - e - k
      infectious_time <- function_data$tPRO[j] - k
      removed_time <- function_data$tERU[j] + d
      
      # Count if individual was EXPOSED on day i
      
      if (exposure_time <= days[i] && days[i] <= infectious_time){
        X = X + 1
        S = S + 1
      }
      
      # Count if individual was INFECTIOUS on day i
      
      if (infectious_time <= days[i] && days[i] <= removed_time){
        Y = Y + 1
      }
      
      # Count if individual was REMOVED on day i
      
      if (days[i] >= removed_time){
        Z = Z + 1
      }
    }
    obs.E.SEIR = c(obs.E.SEIR, X)
    obs.I.SEIR = c(obs.I.SEIR, Y)
    obs.R.SEIR = c(obs.R.SEIR, Z)
    obs.E_added = c(obs.E_added, S)
  }
  
  # Susceptibles over time
  
  obs.S.SEIR <- n - obs.E.SEIR - obs.I.SEIR - obs.R.SEIR
  
  # SEIR graph
  
  if (SEIR.plot == TRUE){
  
    plot(obs.S.SEIR, type = "l", col = "green", xlab = "Time (days)", ylab = "Number of individuals", main = "Infectious states of individuals over time")
    lines(obs.E.SEIR, col = "orange")
    lines(obs.I.SEIR, col = "red")
    lines(obs.R.SEIR, col = "blue")
    abline(h=0, lty = 2)
    abline(h=n, lty = 2)
    grid()
    legend(70, 125, legend=c("susceptible", "exposed", "infectious", "removed"), col = c("green", "orange", "red", "blue"), lty = c(1, 1), cex = 0.9)
  }
  
  return(cbind(obs.S.SEIR, obs.E.SEIR, obs.I.SEIR, obs.R.SEIR, obs.E_added))
}

obs.SEIR_plot(hagelloch, 9, 1, 3, SEIR.plot = TRUE)
```

## Location-based infections

We now display the number of inferctions and relative distance on a plot.

```{r location-based infections, fig.height = 9, echo=FALSE}

plot(hagelloch.df, xlab="x [m]", ylab="y [m]", pch=19, axes=TRUE, cex=sqrt(multiplicity(hagelloch.df)), xlim = c(100, 200), main = "Number of infections in households")

for (i in unique(hagelloch$HN)){
  data_house = hagelloch[hagelloch$HN == i, ]
  text(data_house$x.loc[1]-10, data_house$y.loc[1], i)
  text(25, 100, "Sizes of circles represents number of infections per household")
}
```

## Infectious period histogram

Histogram of infectious periods (including those for individuals who died before recovery).

```{r exploratory 3, echo=FALSE}
infectious_period = (hagelloch$tERU + 3) - (hagelloch$tPRO - 1)
hist(infectious_period, breaks = 10, main = "Histogram of infectious period lengths", xlab = "Time (days)")
```

The most common infectious period can be seen to be around 8 days.